---
layout: post
title: "Pedestrian-Involved Fatal Crashes and Walkability"
description: Using R, I try to ascertain the relationship between road safety and walkability.
date: "2024-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction
## Why, What, and How

Walkability advocates constructed **Walk Score**, available at walkscore.com, as a measurement of the quality of pedestrian-centered infrastructure in cities. Walk Score achieves this by integrating data points like walking times to desired amenities, population and intersection density, and the length of each city block. While Walk Score reflects the convenience of the built environment for walkers and rollers, the same factors are intertwined with implications for road safety. When urban planners make pedestrian-convenient city streets and sidewalks, this usually includes features that prevent lethal collisions with motor vehicles. Additionally, drivers may adjust their behavior to reflect increased foot traffic. Using data from the 2022 U.S. National Highway Traffic Safety Administration’s **Fatality Analysis Reporting System**, the relationship between the conveniences represented by Walk Score and fatal collisions between pedestrians and drivers can be measured and understood.



## Fatal Crashes Involving Pedestrians
Here, the `accident` csv file detailing data for all fatal car accidents in the U.S. was filtered to include only cases appearing within the `pbtype` file, which focuses on details pertaining specifically to accidents between pedestrians and automobiles. Both came from the 2022 NTHSA FARS database (https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars). `pbtype` was first filtered via the `PEDCTYPE` column to remove incidents not involving true pedestrians (i.e., cyclists).
```{r}
require(tidyverse)
```

```{r, echo=FALSE}

accident <- read_csv('/Users/jennenwosu/Documents/individual Research/crashingwalking/FARS2022NationalCSV/accident.csv')
pbtype <- read_csv('/Users/jennenwosu/Documents/individual Research/crashingwalking/FARS2022NationalCSV/pbtype.csv')

```
```{r}
ped_crash <- pbtype %>%
  filter(PEDCTYPE!=0)
ped_crash_acc <- accident %>%
  filter(ST_CASE %in% ped_crash$ST_CASE)
```

## Creating a Clean Data Frame Incorporating Walk Score and City Data with FARS data
The "US Cities Urban Connectivity" dataset by VivianEllis on Kaggle (doi.org/10.34740/kaggle/dsv/7816051) includes information on 100 cities, all of which have populations over 200,000, and their convenience for publicly accessible transportation option.  Its inclusion of Walk Score and 2022 population numbers are useful for this project. To merge this data with the FARS data, some cleaning and reformatting was necessary.

```{r echo=FALSE}
urban_connectivity <- read_csv('/Users/jennenwosu/Documents/individual Research/crashingwalking/urban_connectivity.csv')
```

```{r}
require(usdata)

# Trimming nonrelevant info + matching records
most_pop_cities <- urban_connectivity %>%
  select(City, State, Population_2022_Census, Place_name, 'Walk Score') %>%
  mutate(City = replace(City, City == "Washington D.C.", "Washington")) %>%
  mutate(City = replace(City, City == "Boise City", "Boise")) %>%
  mutate(City = replace(City, City == "St. Petersburg", "Saint Petersburg")) %>%
    mutate(City = replace(City, City == "Lexington", "Lexington-Fayette")) %>%
  mutate(Place_name = replace(Place_name, Place_name == "St. Petersburg, FL", "Saint Petersburg, FL")) %>%
   mutate(Place_name = replace(Place_name, Place_name == "Irvine, CA", "East Irvine, CA")) %>%
  mutate(Place_name = replace(Place_name, Place_name == "Charlotte/Mecklenburg, NC", "Charlotte, NC"))%>%
  mutate(Place_name = replace(Place_name, Place_name == "Nashville/Davidson, TN", "Nashville, TN")) %>%
   mutate(Place_name = replace(Place_name, Place_name == "Lexington/Fayette, KY", "Lexington-Fayette, KY"))%>%
  arrange(-Population_2022_Census) %>%
  head(102) %>%
  rename(WalkScore = "Walk Score")

# Selecting cities appearing within both datasets

ped_crash_acc <- ped_crash_acc %>%
  filter((state2abbr(STATENAME) %in% most_pop_cities$State & tolower(CITYNAME) %in% tolower(most_pop_cities$City))) %>%
  mutate(STATEAB = state2abbr(STATENAME)) %>%
  mutate(CITYNAME = replace(CITYNAME, CITYNAME == "East Irvine", "Irvine"))
  
# Counting up incidents for each city

ped_crash_count <- ped_crash_acc %>%
  mutate(CITYNAME = str_to_title(CITYNAME)) %>%
  mutate(PLACENAME = paste(CITYNAME, STATEAB, sep = ", ")) %>%
  group_by(STATENAME, CITYNAME, STATEAB, PLACENAME) %>%
  summarize(case_count = n()) 
  
# Final step to consolidate Ellis and NHTSA datasets

crash_and_pop <- left_join(ped_crash_count, most_pop_cities, join_by(PLACENAME == Place_name, STATEAB == State, CITYNAME == City))

# Banishing duplicates and erroneously selected cities
crash_and_pop <- crash_and_pop %>%
  drop_na(WalkScore)%>%
  group_by(CITYNAME) %>% 
  summarise_each(funs(first(.[!is.na(.)])))

# Crashes per capita 
crash_and_pop <- crash_and_pop %>%
  mutate(per_capita_crash = case_count / Population_2022_Census)
  

```
## Visualizations of Relevant Relationships
These plots explain both the interest in investigating the relationship between Walk Score and pedestrian-involved fatal collisions, but also the rationale for accounting for population when comparing cities.
```{r}
require(plotly)
# Top Cities for Crashes
crash_and_pop %>%
  arrange(-case_count)%>%
  head(10) %>%
  plot_ly( x = ~reorder(PLACENAME, -case_count), y = ~case_count) %>%
  add_trace(type = 'bar') %>%
  layout(title = "U.S. Cities By Fatal Pedestrian-Involved Car Accidents", xaxis = list(title = "City"), yaxis = list(title = "Number of Crashes"))


# Top Cities for Crashes Per 100,000 people 
crash_and_pop %>%
  arrange(-per_capita_crash)%>%
  head(10) %>%
   plot_ly( x = ~reorder(PLACENAME,  -per_capita_crash), y = ~per_capita_crash*100000) %>%
  add_trace(type = 'bar') %>%
  layout(title = "U.S. Cities By Fatal Pedestrian-Involved Car Accidents, \n Per 100,000 Inhabitants", xaxis = list(title = "City"), yaxis = list(title = "Crashes Per 100,000 (2022 Population)"))
 
# Top Cities for Walk Score
crash_and_pop %>%
  arrange(-WalkScore)%>%
  head(10) %>%
   plot_ly( x = ~reorder(PLACENAME,  -WalkScore), y = ~WalkScore) %>%
  add_trace(type = 'bar') %>%
  layout(title = "10 U.S. Cities With Highest Walk Score", xaxis = list(title = "City"), yaxis = list(title = "Walk Score"))
  
# Bottom Cities for Walk Score
crash_and_pop %>%
  arrange(WalkScore)%>%
  head(10) %>%
  plot_ly( x = ~reorder(PLACENAME,  -WalkScore), y = ~WalkScore) %>%
   add_trace(type = 'bar') %>%
  layout(title = "10 U.S. Cities With Lowest Walk Score \n Out of 100 With Pop. Over 200K", xaxis = list(title = "City"), yaxis = list(title = "Walk Score"))

# Population vs Crashes
scatter1 <- lm(case_count ~ log(Population_2022_Census), crash_and_pop)
summary(scatter1)
crash_and_pop %>%
  plot_ly(x = ~Population_2022_Census, y = ~case_count, hoverinfo = 'text', text = ~paste(PLACENAME), type = 'scatter', mode = 'markers', name = "City") %>%
  layout(title = "Population vs Fatal Pedestrian-Involved Car Accident Incidence", xaxis = list(title = "Population", type = "log"), yaxis = list(title = "Number of Crashes"))%>%
  add_trace(x = ~Population_2022_Census, y = fitted(scatter1), mode = 'lines', hoverinfo = 'skip', name = 'Regression Line')


# Population vs Walk Score
scatter2 <- lm(WalkScore ~ log(Population_2022_Census), crash_and_pop)
summary(scatter2)
crash_and_pop %>%
  plot_ly(x = ~Population_2022_Census, y = ~WalkScore, hoverinfo = 'text', text = ~paste(PLACENAME), type = 'scatter', mode = 'markers', name = "City") %>%
  layout(title = "Population vs Walk Score", xaxis = list(title = "Population", type = "log"), yaxis = list(title = "Walk Score"))%>%
  add_trace(x = ~Population_2022_Census, y = fitted(scatter2), mode = 'lines', hoverinfo = 'skip', name = 'Regression Line')




```

## Regression For Crash Incidence and Walk Score

The method for tackling the guiding question about the relationship between fatal pedestrian-involved crashes and Walk Score was a simple linear regression to check for correlation. The linear model underwent 1000-fold cross-validation with a 70-30 split, from which the resulting average RMSE was about 16.

```{r}
reg_pedcrash <- lm(WalkScore ~ per_capita_crash, crash_and_pop)

crash_and_pop %>%
  plot_ly(x = ~per_capita_crash, y = ~WalkScore, hoverinfo = 'text', text = ~paste(PLACENAME), type = 'scatter', mode = 'markers', name = "City") %>%
  layout(title = "Crashes Per Capita vs Walk Score \n From 100 Cities With Pop. Over 200K", xaxis = list(title = "Fatal Crashes Involving Pedestrians Divided by Population"), yaxis = list(title = "Walk Score"))%>%
  add_trace(x = ~per_capita_crash, y = fitted(reg_pedcrash), mode = 'lines', hoverinfo = 'skip', name = 'Regression Line')

# High p-value!

summary(reg_pedcrash)


set.seed(196)
cvResult <- numeric(1000)
for(i in 1:1000) {

  inds <- sample(x = 1:nrow(crash_and_pop), size = round(.7*nrow(crash_and_pop)),  replace = F) 
  train <- crash_and_pop %>% slice(inds) 

  reg_score_CV <- lm(formula = WalkScore ~ per_capita_crash, 
               data = train)
  
  test <- crash_and_pop %>% slice(-inds)
  
  
  test$preds <- predict(reg_score_CV, newdata = test)
  
  e <- test$WalkScore - test$preds
  se <- e^2
  mse <- mean(se)
  rmse <- sqrt(mse)
  cvResult[i] <- rmse
  

}

mean(cvResult)


```

## Evaluating Alternative Subset of Data

The average RMSE after cross-validation was too high to be confident in pedestrian-involved fatal crashes being a good predictor of walkability as measured by Walk Score, especially with a relatively high p-value of 0.268. Earlier, all incidents coded in FARS as being pedestrian-involved were included. To test whether selecting specific scenarios would eliminate noise and make the relationship clearer, below are the results when restricted to only the accidents that are most obviously relevant to said relationship.

```{r}
# Selecting common pedestrian situations. Chose "walking in the road against traffic" but not "with traffic" due to the former usually being the legal protocol when there are no sidewalks. Refer to FARS/CRSS Pedestrian Bicyclist Crash Typing Manual for PEDCTYPE code meanings.

ped_crash_small <- pbtype %>%
  filter(PEDCTYPE== 341 | PEDCTYPE==342 | PEDCTYPE==430 | PEDCTYPE==440 | PEDCTYPE==510 | PEDCTYPE == 520 | PEDCTYPE==590 | PEDCTYPE==710 | PEDCTYPE==720 | PEDCTYPE==770 | PEDCTYPE==781 | PEDCTYPE==782 | PEDCTYPE==791 | PEDCTYPE==792 | PEDCTYPE==795 | PEDCTYPE==794 | PEDCTYPE==799 )
ped_crash_acc_small <- accident %>%
  filter(ST_CASE %in% ped_crash_small$ST_CASE)

ped_crash_acc_small <- ped_crash_acc_small %>%
  filter((state2abbr(STATENAME) %in% most_pop_cities$State & tolower(CITYNAME) %in% tolower(most_pop_cities$City))) %>%
  mutate(STATEAB = state2abbr(STATENAME))

ped_crash_count_small <- ped_crash_acc_small %>%
  mutate(CITYNAME = str_to_title(CITYNAME)) %>%
  mutate(PLACENAME = paste(CITYNAME, STATEAB, sep = ", ")) %>%
  group_by(STATENAME, CITYNAME, STATEAB, PLACENAME) %>%
  summarize(case_count = n()) 

crash_and_pop_small <- left_join(ped_crash_count_small, most_pop_cities, join_by(PLACENAME == Place_name, STATEAB == State, CITYNAME == City))

# Banishing duplicates and erroneously selected cities
crash_and_pop_small <- crash_and_pop_small %>%
  drop_na(WalkScore)%>%
  group_by(CITYNAME) %>% 
  summarise_each(funs(first(.[!is.na(.)])))

crash_and_pop_small <- crash_and_pop_small %>%
  mutate(per_capita_crash = case_count / Population_2022_Census)

# MUCH smaller sample size - will this cause problems? Let's see.


```
## Re-running Regression on Smaller Subset

```{r}
reg_pedcrash_small <- lm(WalkScore ~ per_capita_crash, crash_and_pop_small)

crash_and_pop_small %>%
  plot_ly(x = ~per_capita_crash, y = ~WalkScore, hoverinfo = 'text', text = ~paste(PLACENAME), type = 'scatter', mode = 'markers', name = "City") %>%
  layout(title = "Crashes Per Capita vs Walk Score \n Subsetted from 100 Cities With Pop. Over 200K", xaxis = list(title = "Fatal Crashes Involving Pedestrians (selected PEDCTYPE codes) Divided by Population"), yaxis = list(title = "Walk Score"))%>%
  add_trace(x = ~per_capita_crash, y = fitted(reg_pedcrash_small), mode = 'lines', hoverinfo = 'skip', name = 'Regression Line')


summary(reg_pedcrash_small)
# Only a tiny change in p-value...

set.seed(196)
cvResult2 <- numeric(1000)
for(i in 1:1000) {

  inds <- sample(x = 1:nrow(crash_and_pop_small), size = round(.7*nrow(crash_and_pop_small)),  replace = F) 
  train <- crash_and_pop_small %>% slice(inds) 

  reg_score_CV <- lm(formula = WalkScore ~ per_capita_crash, 
               data = train)
  
  test <- crash_and_pop_small %>% slice(-inds)
  
  
  test$preds <- predict(reg_score_CV, newdata = test)
  
  e <- test$WalkScore - test$preds
  se <- e^2
  mse <- mean(se)
  rmse <- sqrt(mse)
  cvResult2[i] <- rmse
  

}

mean(cvResult2)

```
## Conclusion

Regression analysis did not reveal a statistically significant relationship between fatal pedestrian-involved crash incidence and Walk Score. Predictions made with a simple linear model for these two variables had considerable error. Pruning cases for relevance did not seem to have any effect on this error or the statistical significance of the relationship. The inconclusive finding may be due to the fact that this work included only a narrow range of cities. Additionally, perhaps Walk Score, intended for use by prospective tenants and homeowners to choose where to live, is not the best performing measure for walkability in the context of road safety. More work would be needed for a definitive answer to the question driving this project.
